---
title: "Central Limit Theorem"
author: "Heli Pykälä"
date: '2016-05-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this report we explore Central Limit Theorem by calculating averages of samples from exponential distribution.

## Simulation

First we set parameters for the simulation. Exponential distribution has one parameter: rate marked by $\lambda$. In this simulation we use $\lambda = 0.2$. We draw 40 samples from $Exp(\lambda)$ and repeat the simulation 1000 times. Call the number of samples and the number of simulations $n$ and $m$ respectively.
```{r params}
library(ggplot2)
lambda <- .2
n <- 40
m <- 1000
```

Then set seed for reproducibility and draw $n*m$ samples from the distribution. 
```{r simulation}
set.seed(135)
samples <- rexp(n*m, lambda)
```


##The simulated exponential distribution

Theoretical mean of exponential distribution is $\mu = 1/\lambda$ and it's standard deviation is $\sigma = 1/\lambda$


```{r}
1/lambda #theoretical mean and sd
mean(samples) #mean of the sample
sd(samples) #standard deviation of the sample
```

The mean and standard deviation of the samples follow closely to the theretical values.

Then Plot the density histogram of the samples with the theoretical density given $\lambda = 0.2$. Theoretical mean is the thick black vertical line and calculated mean of the sample is the cyan vertical line.
```{r}
library(ggplot2)
dfSamples <- data.frame(x = samples)
ggplot(dfSamples,aes(x=x))+
    geom_histogram(colour = "black", fill = "steelblue", binwidth = 2, boundary = 0,  aes(y = ..density..)) +
    stat_function(fun = dexp, size = 2, args = list(rate = lambda)) +
    geom_vline(xintercept = 1/lambda, size = 2, color = "black") +
    geom_vline(xintercept = mean(samples), size = 1, color =  "cyan")
```

The distribution looks very similar to the theoretical density function of the exponential distribution.

## Averages of exponential samples

Arrange the samples into a matrix with 1000 rows and 40 columns and calculate columnwise means to acquire 1000 samples of averages of 40 exponentials.

```{r}
mat <- matrix(samples, nrow = m, ncol = n)
means <- apply(mat, 1, mean)

```

Theoretically the mean of the means should follow a normal distribution centered at the population mean, $\mu = 1/\lambda$ and the standard deviation should be $\frac{\sigma}{\sqrt{n}} = \frac{1}{\lambda\sqrt{n}}$

```{r}
1/lambda #theoretical mean
mean(means) #mean of the sample
1/lambda/sqrt(n) #theoretical sd
sd(means) #standard deviation of the sample
```

The mean and standard deviation of the sample of 1000 averages of 40 exponentials follow quite closely to the theoretical values.

Then plot the density histogram of the averages along with the theoretical normal distribution with $\mu = 1/\lambda$ and $sd = \frac{1}{\lambda\sqrt{n}}$. 
```{r}
dfMeans <- data.frame(x = means)
ggplot(dfMeans,aes(x=x))+
    geom_histogram(colour = "black", fill = "steelblue", binwidth = .25, aes(y = ..density..)) + 
    stat_function(fun = dnorm, size = 2, args = list(mean = 1/lambda, sd =  1/lambda/sqrt(n))) +
    geom_vline(xintercept = 1/lambda, size = 2, color = "black") +
    geom_vline(xintercept = mean(means), size = 1, color = "cyan")
```


