---
title: "Central Limit Theorem"
author: "Heli Pykälä"
date: '2016-05-10'
output: 
  html_document: 
    fig_height: 3
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The Central Limit Theorem states that averages of independent and indentically distributed random variables should follow the standard normal distribution when properly normalized. We test this by simulation. We find that the Central Limit Theorem seems to hold for averages of 40 samples from the exponential distribution $Exp(0.2)$

## Simulation

First we set parameters for the simulation. Exponential distribution has one parameter: rate marked by $\lambda$. In this simulation we use $\lambda = 0.2$. We draw 40 samples from $Exp(\lambda)$ and repeat the simulation 1000 times. Call the number of samples and the number of simulations $n$ and $m$ respectively.
```{r params}
library(ggplot2)
lambda <- .2
n <- 40
m <- 1000
```

Then set seed for reproducibility and draw $n*m$ samples from the distribution. 
```{r simulation}
set.seed(135)
samples <- rexp(n*m, lambda)
```


##The simulated exponential distribution

First we study a sample of 1000 exponentials:

```{r}
s <- samples[1:1000]
```
Theoretical mean of exponential distribution is $\mu = \frac{1}{\lambda}$ and it's standard deviation is $\sigma = \frac{1}{\lambda}$

```{r}
1/lambda #theoretical mean and sd
mean(s) #mean of the sample
sd(s) #standard deviation of the sample
```

The mean and standard deviation of the samples are close to the theoretical values.

Then Plot the density histogram of the samples with the theoretical density given $\lambda = 0.2$. Theoretical mean is the thick black vertical line and calculated mean of the sample is the cyan vertical line.
```{r}
library(ggplot2)
dfSamples <- data.frame(x = s)
ggplot(dfSamples,aes(x=x))+
    geom_histogram(colour = "black", fill = "steelblue", binwidth = 2, boundary = 0,  aes(y = ..density..)) +
    stat_function(fun = dexp, size = 2, args = list(rate = lambda)) +
    geom_vline(xintercept = 1/lambda, size = 2, color = "black") +
    geom_vline(xintercept = mean(samples), size = 1, color =  "cyan")
```

The distribution looks very similar to the theoretical density function of the exponential distribution. We can note that this original distribution has very different characteristics than the normal distribution. It is very skewed, always > 0 and has the largest density at 0.

## The sample mean

Arrange the samples into a matrix with 1000 rows and 40 columns and calculate column-wise means to acquire 1000 samples of means of 40 exponentials.

```{r}
mat <- matrix(samples, nrow = m, ncol = n)
means <- apply(mat, 1, mean)

```
### Sample Mean versus Theoretical Mean

Theoretically the mean of the means should be the population mean, $\mu = \frac{1}{\lambda}$

```{r}
1/lambda #theoretical mean
mean(means) #mean of the sample
```
The sample mean is close to the theoretical value.

### Sample Variance versus Theoretical Variance

The standard deviation should be $\frac{\sigma}{\sqrt{n}} = \frac{1}{\lambda\sqrt{n}}$

```{r}
1/lambda/sqrt(n) #theoretical standard deviation
sd(means) #standard deviation of the sample
```
The standard deviation of the sample is close to the theoretical value.

### Distribution

Normalize the values by subtracting the mean and dividing by the standard deviation

```{r}
scaledMeans <- scale(means, center = 1/lambda, scale = 1/lambda/sqrt(n))
```

Plot the density histogram of the scaled averages along with the standard normal distribution. 
```{r}
dfMeans <- data.frame(x = scaledMeans)
ggplot(dfMeans,aes(x=x))+
    geom_histogram(colour = "black", fill = "steelblue", binwidth = .25, aes(y = ..density..)) + 
    stat_function(fun = dnorm, size = 2) +
    geom_vline(xintercept = 0, size = 2, color = "black") +
    geom_vline(xintercept = mean(scaledMeans), size = 1, color = "cyan")
```

The histogram looks similar to the theoretical normal distribution.

##Conclusion

The Central Limit Theorem seems to hold for averages of 40 exponentials. The original exponential distribution is really different from the normal distribution, but when we calculate the average of 40 random samples the resulting random variable seems to follow the normal distribution with expected parameters reasonably well.
